---
title: "PCA Rotations: What are they"
author: "Sam Passmore"
editor: visual
category: methods
---

I have been asked a couple of times about the use of rotations in Principle Component Analysis (PCA).
Each time it is far enough apart that I forget what I said the previous time, so I have written some notes here, with a particular focus on Oblique Minimum rotation.

# What happens in a PCA

The primary use I have seen PCA is to understand latent variation within a dataset.
That is, when someone has research goal to measure something that is not directly measurable, one way around that is to ask a number of measurable questions and understand how they correlate to get a measure of that underlying factor.
A principle component analysis identifies what variation is shared amongst theose variables, and loads their variation onto a single component, which is often inferred to be an indreicted measurement of that latent (or immeasurable) variable.
Equally, in an observed dataset (i.e. a datatset where there was no opportunity to design the questions or structure of the data *a-priori*), it may be the case that there are some set of variables that you might predict 'hang-together' because of some underlying, but not directly measured reason.

This is conceptually why I think people use PCA, but what is actually happening to the numbers?
First, I will look at this using a simulated dataset, than we can look at how this will work with a real linguistic datatset.

Here is the set up for the simulated data:

Numerically, PCA is derived from a correlation matrix.
That is, for each variable you are interested in, calculate how they are correlated to each other and put them into a square matri.

Assuming that we have a standardized (the mean is zero for all variables) Principle component analysis are derived from the covariance of the variables in the matrix.
Let's take a small example, using data from the R package [languageR](https://cran.r-project.org/web/packages/languageR/languageR.pdf).
We are going to use the datatset '*english*'.
This data set gives mean visual lexical decision latencies and word naming latencies to 2284 monomorphemic English nouns and verbs, averaged for old and young subjects, with various pre- dictor variables.
For each word, there is an average latency value for an Old and Young participant.
I choose this datatset only because I see that it has been used for PCA before.
Specifically the variables

```{r}
library(languageR)

data(english)
head(english)

items = english[english$AgeSubject == "young", c(18:27)]

items.pca = prcomp(items, center = TRUE, scale = TRUE)
x = as.data.frame(items.pca$rotation[,1:4])
items$PC1 = items.pca$x[,1]
items$PC2 = items.pca$x[,2]
items$PC3 = items.pca$x[,3]
items$PC4 = items.pca$x[,4]
items2 = english[english$AgeSubject != "young", ]
items2$PC1 = items.pca$x[,1]
items2$PC2 = items.pca$x[,2]
items2$PC3 = items.pca$x[,3]
items2$PC4 = items.pca$x[,4]
english = rbind(items, items2)
```

```{r}
library(psych)

# --- Step 1: Create example data with clear 2-factor structure ---
set.seed(123)
n <- 200
F1 <- rnorm(n)
F2 <- rnorm(n)

# Variables mainly loading on one factor
V1 <- 0.8*F1 + 0.1*F2 + rnorm(n, 0, 0.2)
V2 <- 0.7*F1 + 0.2*F2 + rnorm(n, 0, 0.2)
V3 <- 0.1*F1 + 0.8*F2 + rnorm(n, 0, 0.2)
V4 <- 0.2*F1 + 0.7*F2 + rnorm(n, 0, 0.2)

df <- data.frame(V1, V2, V3, V4)

# --- Step 2: Compute correlation matrix ---
R <- cor(df)

# --- Step 3: Factor analysis with 2 factors ---
fa_unrotated <- fa(df, nfactors = 2, rotate = "none")
print(fa_unrotated$loadings, cutoff = 0)

# --- Step 4: Varimax rotation (orthogonal) ---
fa_varimax <- fa(df, nfactors = 2, rotate = "varimax")
print(fa_varimax$loadings, cutoff = 0)

# --- Step 5: Promax rotation (oblique minimum rotation) ---
fa_promax <- fa(df, nfactors = 2, rotate = "promax")
print(fa_promax$loadings, cutoff = 0)

plot(fa_unrotated$scores[,1], fa_unrotated$scores[,2])
cor(fa_unrotated$scores[,1], fa_unrotated$scores[,2])

plot(fa_varimax$scores[,1], fa_varimax$scores[,2])
cor(fa_varimax$scores[,1], fa_varimax$scores[,2])

plot(fa_promax$scores[,1], fa_promax$scores[,2])
cor(fa_promax$scores[,1], fa_promax$scores[,2])


```
